Mammalian protein-coding genes exhibit widespread beneficial mutations that are not adaptive

### 2024-03-04 manuscript/main-PNAS.tex
Manuscript revision 2

% Please add here a significance statement to explain the relevance of your work
\significancestatement{
       The extent to which adaptation to changing environments is shaping genomes is a central question in molecular evolution.
       To quantify the rate of adaptation, population geneticists have typically used signatures of positive selection.
       However, mutations restoring an ancestral sequence of higher fitness lost by genetic drift are also positively selected, but they do not respond to a change in the environment.
       In this study, we have managed to distinguish beneficial mutations that are due to changing environments and those that are restoring pre-existing functions in mammals.
       We show that a substantial proportion of beneficial mutations cannot be interpreted as adaptive.
}

@@

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Adaptation $|$ Back-mutations $|$ Phylogenetics $|$ Population-genetics $|$ Codon models }

\begin{abstract}
Mutations can be beneficial by bringing innovation to their bearer, allowing them to adapt to environmental change.
These mutations are typically unpredictable since they respond to an unforeseen change in the environment.
However, mutations can also be beneficial because they are simply restoring an ancestral sequence of higher fitness that was lost due to genetic drift.
In contrast to adaptive mutations, these beneficial back-mutations can be predicted if the underlying fitness landscape is stable and known.
The contribution of such beneficial back-mutations to molecular evolution has been widely neglected mainly because their detection is very challenging.
We have here reconstructed protein-coding-gene fitness landscapes shared between mammals, using mutation-selection models and a multi-species alignments across 87 mammals.
These fitness landscapes have allowed us to predict the fitness effect of polymorphisms found in 28 mammalian populations.
Using methods that quantify selection at the population level, we have confirmed that beneficial back-mutations are indeed positively selected in extant populations.
Our work confirms that deleterious substitutions are accumulating in mammals and are being reverted, generating a balance in which genomes are damaged and restored simultaneously at different loci.
We observe that beneficial back-mutations represent between 15\% and 45\% of all beneficial mutations in 24 of 28 populations analyzed, showing that a substantial part of ongoing positive selection is not driven by adaptation to environmental change in mammals.
\end{abstract}

@@

\firstpage[3]{3}
\dropcap{A}daptation is one of the main processes shaping the diversity of forms and functions across the tree of life~\cite{darwin_origin_1859}.
Evolutionary adaptation is tightly linked to environmental change and species responding to this change~\cite{merrell_adaptive_1994, gavrilets_adaptive_2009}.
For adaptation to occur, there must be variation within populations, which mostly appears via mutations in the DNA sequence.
While neutral mutations will not impact an individual's fitness, deleterious mutations have a negative effect, and beneficial mutations improve their bearer's fitness.
A beneficial mutation is thus more likely than a neutral mutation to invade the population and reach fixation, resulting in a substitution at the species level.
Upon environmental change, because adaptive beneficial mutations toward new fitness optima are more likely, the number of substitutions also increases (fig.~\ref{fig:fitness-landscape}A).
An increased substitution rate is thus commonly interpreted as a sign of adaptation~\cite{mcdonald_adaptative_1991, smith_adaptive_2002, welch_estimating_2006}.
The availability of large-scale genomic data and the development of theoretical models have enabled the detection and quantification of substitution rate changes across genes and lineages~\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019}.
These approaches, now common practice in evolutionary biology, have helped better understand the processes underpinning the rates of molecular evolution, contributing to disentangling the effects of mutation, selection and drift in evolution~\cite{lynch_mutation_2023}.
However, a collateral effect has been conflating beneficial mutations with adaptive evolution when adaptive evolution is not the only process that can lead to beneficial mutations~\cite{charlesworth_other_2007, mustonen_fitness_2009, jones_shifting_2017}.

In a constant environment, a deleterious mutation can reach fixation by genetic drift~\cite{Ohta1992}.
A new mutation restoring the ancestral fitness will thus be beneficial (fig.~\ref{fig:fitness-landscape}B), even though the environment has not changed~\cite{gillespie_ohta_1995, hartl_compensatory_1996, sella_application_2005, mustonen_fitness_2009, cvijovic_fate_2015}.
The restoration of the ancestral fitness can either happen through a mutation at a different locus –~called a compensatory mutation~\cite{hartl_compensatory_1996, mustonen_fitness_2009}, or at the locus of the initial mutation –~called a beneficial back-mutation~\cite{piganeau_estimating_2003, charlesworth_other_2007}.
While compensatory mutations change the sequence and thus induce genetic diversification, beneficial back-mutations reduce genetic diversity and do not contribute to genetic innovation.

Although Tomoko Ohta considered beneficial back-mutations negligible in her nearly-neutral theory~\cite{Ohta1992}, their importance has now been acknowledged for
expanding populations~\cite{charlesworth_other_2007}.
However, differentiating between an adaptive mutation and a beneficial back-mutation remains challenging~\cite{chi_detecting_2020}.
Indeed, an adaptive mutation responding to a change in the environment and a beneficial back-mutation have equivalent fitness consequences for their bearer~\cite{charlesworth_other_2007}.
Similarly, at the population level, both types of mutations will result in a positive transmission bias of the beneficial allele.
However, at the macro-evolutionary scale, the consequences of these two types of mutations are fundamentally different.
While adaptive mutations promote phenotype diversification (fig.~\ref{fig:fitness-landscape}C), beneficial back-mutations promote phenotype stability and may help preserve well-established biological systems (fig.~\ref{fig:fitness-landscape}D).
Additionally, the direction of adaptive evolution is unpredictable because it is caused by an unforeseen change in the environment and, hence, in the underlying fitness landscape~\cite{bazykin_changing_2015}.
On the other hand, beneficial back-mutations are predictable because, under a stable fitness landscape, any change from non-optimal to optimal amino acids will move back the site towards the equilibrium expected under the fitness landscape~\cite{moses_inferring_2009, fischer_germline_2011, chen_hunting_2021}.
It can then be distinguished from truly novel beneficial mutations because the latter are not expected to mutate mutate towards the amino acids of higher fitnesses defined by the stable fitness landscape but rather mutate to amino acids showing a diversified pattern (fig.~\ref{fig:fitness-landscape}).

\begin{figure*}[htb]
       \centering
       \includegraphics[width=17.8cm, page=1]{figure1}
       \caption{
              (A \& B) For a given codon position of a protein-coding DNA sequence, amino acids (x-axis) have different fitness values (y-axis).
              Under a fluctuating fitness landscape (A), these fitnesses change with time.
              The protein sequence follows the moving target defined by the amino-acid fitnesses. Since substitutions are preferentially accepted if they are in the direction of this target, substitutions are, on average, adaptive.
              At the phylogenetic scale (C), beneficial substitutions are common (positive signs), promoting phenotype diversification across species.
              Under a stable fitness landscape (B), most mutations reaching fixation are either slightly deleterious reaching fixation due to drift or are beneficial back-mutations restoring a more optimal amino acid.
              At the phylogenetic scale (D), deleterious substitutions (negative signs) are often reverted via beneficial back-mutations (positive signs), promoting phenotype stability and preserving well-established biological systems.
              Even though, individually, any back-mutation might have a small beneficial effect on its bearer, we expect beneficial back-mutations to be scattered across the genome and the genome-wide signature of beneficial back-mutations to be detectable and quantifiable.}
       \label{fig:fitness-landscape}
\end{figure*}

\begin{figure*}[htb]
       \centering
       \includegraphics[width=17.8cm, page=1]{figure2}
       \caption{
              Selection coefficients at the phylogenetic and population scales.
              At the phylogenetic scale (A), we estimated the amino-acid fitness for each site from protein-coding DNA alignments using mutation-selection codon models.
              For every possible mutation, the difference in amino-acid fitness before and after the mutation allows us to compute the selection coefficient at the phylogenetic scale ($\Sphy$).
              Depending on $\Sphy$ (B) mutations can be predicted as deleterious ($\SphyDel$), nearly-neutral ($\SphyNeu$) or beneficial back-mutations ($\SphyBen$) toward a fitter amino acid and repairing existing functions.
              At the population scale, each observed single nucleotide polymorphism (SNP) segregating in the population can also be classified according to its $\Sphy$ value (C).
              Occurrence and frequency in the population of non-synonymous polymorphisms, contrasted to synonymous polymorphisms (deemed neutral) is used to estimate selection coefficients (D-E) at the population scale ($\Spop$), for each class of selection ($\SphyDel$, $\SphyNeu$, $\SphyBen$).
              We can thus assess whether $\Sphy$ predicts $\Spop$ and compute \textit{precision} (F) and \textit{recall} (G) for each class.
              The \textit{recall} value for class $\SphyBen$ is the probability of back-mutations among all beneficial ones (G).
       }
       \label{fig:method}
\end{figure*}

@@

Theoretically, we can thus use large-scale genomic data to assess whether the fitness estimated at the phylogenetic scale predict the fitness effect at the population scale for both deleterious and beneficial back-mutations.
The placental mammals represent an excellent study system to perform such analysis.
Having originated $\sim$102 million years ago, they diversified quickly~\cite{foley_genomic_2023}.
Additionally, polymorphism data are available for many species~\cite{howe_ensembl_2021}, as are high quality protein-coding DNA alignments across the genome~\cite{ranwez_orthomam_2007, scornavacca_orthomam_2019}.
By performing our analysis on 14,509 orthologous protein-coding genes across 87 species, we focus on genes shared across all mammals in our dataset and not newly functionalized genes in a lineage.

For each gene, fitting the mutation-selection model to the multi-species sequence alignment, assuming that the underlying fitness landscape is stable along the phylogenetic tree, allows us to obtain relative fitnesses for all amino acids for each site of the alignment (fig.~\ref{fig:method}A).
Given these relative fitnesses ($F$) for each amino acid, the difference in fitness between a pair of amino acids is the scaled selection coefficient ($\Sphy = \Delta F$), which is formally the product of the selection coefficient at the individual level ($s$) and the effective population size ($\Ne$), as $\Sphy = 4 \Ne \times s$. The value of $\Sphy$ informs us on the strength of selection exerted on amino acids changes.
Thus, according to its $\Sphy$ value, we can classify any mutation as either a deleterious mutation toward a less fit amino acid ($\SphyDel \coloneqq \Sphy < -1$), a nearly-neutral mutation ($\SphyNeu \coloneqq -1 < \Sphy <1$) or a beneficial back-mutation toward a known fitter amino acid ($\SphyBen \coloneqq \Sphy > 1$).

Having identified which potential DNA changes represent beneficial back-mutations (fig.~\ref{fig:method}B), we retrieved polymorphism data from 28 wild and domesticated populations belonging to 6 genera (\textit{Equus}, \textit{Bos}, \textit{Capra}, \textit{Ovis}, \textit{Chlorocebus}, and \textit{Homo}) to assess the presence of beneficial back-mutations at the population scale.
We focused on mutations currently segregating within populations and substitutions in the terminal branches, and checked if any of these observed changes were beneficial back-mutations (fig.~\ref{fig:method}C-E).
A similar approach demonstrated the presence of beneficial back-mutations in humans~\cite{moses_inferring_2009, fischer_germline_2011} and in plants~\cite{chen_hunting_2021}.
However, the model used to reconstruct the static fitness landscape in these studies can only be applied to deeply conserved protein domains in the tree of life, which corresponds to a subpart of the proteome that evolves slowly.
The mutation-selection model used in the present work integrates phylogenetic relationships, and thus allows us to estimate the fitness landscape in shallower phylogenetic trees, and therefore can be applied almost genome-wide~\cite{rodrigue_mutationselection_2010}.

We first quantified the likelihood of any DNA mutation to be a beneficial back-mutation, that is, whenever a DNA mutation increases fitness under a stable fitness landscape ($\SphyBen$).
Subsequently, by quantifying the total amount of beneficial mutations in the current population across all types of DNA mutations ($\SphyDel$, $\SphyDel$ and $\SphyBen$), we could tease apart beneficial back-mutations from beneficial adaptive mutations resulting from a change in the fitness landscape.
Altogether, in this study, by integrating large-scale genomic datasets at both phylogenetic and population scales, we propose a way to explicitly quantify the contribution of beneficial back-mutations to positive selection across the entire exome of the six genera (figs.~\ref{fig:method}F-G).

\section*{Results}\label{sec:results}

\subsection*{Selection along the terminal branches}
First, we assessed whether fitness effects derived from the mutation-selection model at the phylogenetic scale predict selection occurring in terminal branches.
We recovered the mutations that reached fixation in the terminal branches of the six genera.
We only considered mutations fixed in a population as substitutions in the corresponding branch by discarding mutations segregating in our population samples.
We classified each substitution identified in the terminal branches as either $\SphyDel$, $\SphyNeu$, or $\SphyBen$ depending on its $\Sphy$ value obtained at the phylogenetic scale (fig.~\ref{fig:method}A-C).
Importantly, the mammalian alignment used to estimate the amino-acid fitness landscape did not include the genera for which we estimated selection at the population scale as this would bias the estimate of the fitness landscape and generate circular reasoning.
Because $\Sphy$ was based on estimating fitness effects at the phylogenetic scale for all mammals, substitutions with $\Sphy>1$ ($\SphyBen$) bring the population toward an amino acid predicted to be fitter in mammals.
These changes are thus considered as beneficial back-substitutions rather than a species-specific adaptive change.
Among all the substitutions found in each terminal branch, between 10 and 13\% were $\SphyBen$, while instead beneficial back-mutations represent between 0.9 and 1.2\% of all non-synonymous mutations (table~S1 and fig.~\ref{fig:homo-afr-results}A-B for humans).
In principle, beneficial back-mutations are bound to reach fixation more often than neutral mutations.
Hence, we calculated the $\dnds$ ratio of non-synonymous over synonymous divergence for all terminal lineages, focusing on the non-synonymous changes predicted as beneficial back-mutations ($\dn(\SphyBen) / \ds$).
We obtained values between 1.17 and 1.75 in the different lineages (table~S2), implying i) that $\SphyBen$ mutations reached fixation more frequently than synonymous mutations that are supposed to be neutral, and ii) that these back-mutations are effectively beneficial.

This result further indicated that using $\dnds$ as an estimate of adaptation is biased due to the presence of beneficial back-mutations among the non-synonymous substitutions.
By discarding all beneficial back-mutations we can obtain an estimate of $\dnds$ which is not inflated.
By comparing these two ways of calculating $\dnds$ (see section~\ref{subsec:substitution-mapping-in-the-terminal-branch} in Materials \& Methods), we calculated that beneficial back-mutations inflate $\dnds$ values by between 9 and 12\% across genera while only representing between 0.9 and 1.2\% of non-synonymous mutations (table~S3).

@@

In humans, some SNPs have been associated with specific clinical prognosis terms obtained by clinical evaluation of the impact of variants on human Mendelian disorders ~\cite{landrum_clinvar_2018}.
Although this classification also relies on deep protein alignments and therefore cannot be considered an independent result from our own, it does provide a consistency check if the effect of a mutation on human health is in line with its fitness effect predicted by our method.
Therefore, we investigated whether the non-synonymous SNPs classified as $\SphyDel$ or $\SphyBen$ showed enrichment in specific clinical terms compared to SNPs classified as $\SphyNeu$.
Our results show that SNPs predicted as deleterious are associated with clinical terms such as \textit{Likely Pathogenic} and \textit{Pathogenic}, implying that, in general, the selective pressure of a mutation exerted across mammals is also predictive of its clinical effect in humans (table~S4)~\cite{sullivan_leveraging_2023}.
Conversely, back-mutations are associated with clinical terms such as \textit{Benign} and \textit{Likely Benign}, which shows that back-mutations are less likely to be functionally damaging (table~S5).

@@

Mutations predicted as $\SphyNeu$ were effectively composed of a mix of neutral and selected mutations with varying \textit{precision} (36--63\%) and \textit{recall} (32--45\%) across the different populations (table~\ref{table:proba}, fig.~\ref{fig:homo-afr-results}D for humans).
The variable proportions between populations can be explained by the effective number of individuals in the population ($\Ne$), a major driver of selection efficacy.
Moreover, estimates of mutation rate per generation ($u$), from \cite[Bergeron et al]{bergeron_evolution_2023} and \cite[Orlando et al]{orlando_recalibrating_2013}, and Watterson’s $\theta$ obtained from the synonymous SFS as in \cite[Achaz]{achaz_frequency_2009} allow us to obtain $\Ne$ through $\Ne=\theta/(4 \times u)$.
Using correlation analyses that accounted for phylogenetic relationship (see section \ref{subsec:correlation-diversity} in Materials \& Methods), we found that higher $\Ne$ was associated with a smaller proportion of nearly-neutral mutations ($r^2=0.31$, $p=0.001$, fig.~\ref{fig:diversity}A).
This result follows the prediction of the nearly-neutral theory and suggests that in populations with higher diversity (e.g.,~\textit{Bos} or \textit{Ovis}), discrimination between beneficial and deleterious mutations is more likely to occur (figs.~S1-S3).
Conversely, many more mutations are effectively neutral in populations with lower diversity (e.g.,~\textit{Homo}).

>>>>

% Please add here a significance statement to explain the relevance of your work
\significancestatement{
       The extent to which adaptation to changing environments is shaping genomes is a central question in molecular evolution. To quantify the rate of adaptation, population geneticists have typically used signatures of positive selection. However, mutations restoring an ancestral state of higher fitness lost by genetic drift are also positively selected, but they do not respond to a change in the environment. In this study, we have managed to distinguish beneficial mutations that are due to changing environments and those that are restoring pre-existing functions in mammals. We show that a substantial proportion of beneficial mutations cannot be interpreted as adaptive.
}

@@

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Adaptation $|$ Nearly-neutral $|$ Phylogenetics $|$ Population-genetics $|$ Codon models }

\begin{abstract}
       Mutations can be beneficial by bringing innovation to their bearer, allowing them to adapt to environmental change. These mutations are typically unpredictable since they respond to an unforeseen change in the environment. However, mutations can also be beneficial because they are simply restoring a state of higher fitness that was lost due to genetic drift in a stable environment. In contrast to adaptive mutations, these beneficial non-adaptive mutations can be predicted if the underlying fitness landscape is stable and known. The contribution of such non-adaptive mutations to molecular evolution has been widely neglected mainly because their detection is very challenging. We have here reconstructed protein-coding-gene fitness landscapes shared between mammals, using mutation-selection models and a multi-species alignments across 87 mammals. These fitness landscapes have allowed us to predict the fitness effect of polymorphisms found in 28 mammalian populations. Using methods that quantify selection at the population level, we have confirmed that beneficial non-adaptive mutations are indeed positively selected in extant populations. Our work confirms that deleterious substitutions are accumulating in mammals and are being reverted, generating a balance in which genomes are damaged and restored simultaneously at different loci. We observe that beneficial non-adaptive mutations represent between 15\% and 45\% of all beneficial mutations in 24 of 28 populations analyzed, showing that a substantial part of ongoing positive selection is not driven solely by adaptation to environmental change in mammals.
\end{abstract}

@@

\firstpage[1]{2}
\dropcap{A}daptation is one of the main processes shaping the diversity of forms and functions across the tree of life~\cite{darwin_origin_1859}.
Evolutionary adaptation is tightly linked to environmental change and species responding to this change~\cite{merrell_adaptive_1994, gavrilets_adaptive_2009}.
For adaptation to occur, there must be variation within populations, which mostly appears via mutations in the DNA sequence.
While neutral mutations will not impact an individual's fitness, deleterious mutations have a negative effect, and beneficial mutations improve their bearer's fitness.
A beneficial mutation is thus more likely than a neutral mutation to invade the population and reach fixation, resulting in a substitution at the species level.
Upon environmental change, because adaptive beneficial mutations toward new fitness optima are more likely, the number of substitutions also increases (fig.~\ref{fig:fitness-landscape}A).
An increased substitution rate is thus commonly interpreted as a sign of adaptation~\cite{mcdonald_adaptative_1991, smith_adaptive_2002, welch_estimating_2006}.
The availability of large-scale genomic data and the development of theoretical models have enabled the detection and quantification of substitution rate changes across genes and lineages~\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019}.
These approaches, now common practice in evolutionary biology, have helped better understand the processes underpinning the rates of molecular evolution, contributing to disentangling the effects of mutation, selection and drift in evolution~\cite{lynch_mutation_2023}.
However, a collateral effect has been conflating beneficial mutations with adaptive evolution when adaptive evolution is not the only process that can lead to beneficial mutations~\cite{charlesworth_other_2007, mustonen_fitness_2009, jones_shifting_2017}.

\subsection*{Beneficial yet non-adaptive mutations}

In a constant environment, a deleterious mutation can reach fixation by genetic drift~\cite{ohta_nearly_1992}.
A new mutation restoring the ancestral fitness will thus be beneficial (fig.~\ref{fig:fitness-landscape}B), even though the environment has not changed~\cite{gillespie_ohta_1995, hartl_compensatory_1996, sella_application_2005, mustonen_fitness_2009, cvijovic_fate_2015}.
We will refer to as beneficial non-adaptive mutations to those mutations that restore the ancestral fitness under the assumption that the fitness landscape has not changed~\cite{piganeau_estimating_2003, charlesworth_other_2007}.
Such mutations can happen through a mutation at a different locus, in which case it is called a compensatory mutation~\cite{hartl_compensatory_1996, mustonen_fitness_2009}.
While compensatory mutations change the sequence and thus induce genetic diversification, beneficial non-adaptive mutations at the locus of the initial mutation reduce genetic diversity and do not contribute to genetic innovation, which are the focus of this manuscript.
Although Tomoko Ohta considered beneficial non-adaptive mutations negligible in her nearly-neutral theory~\cite{ohta_nearly_1992}, their importance has now been acknowledged for expanding populations~\cite{charlesworth_other_2007}.
However, differentiating between an adaptive mutation and a beneficial non-adaptive mutation remains challenging~\cite{chi_detecting_2020}.
Indeed, an adaptive mutation responding to a change in the environment and a beneficial non-adaptive mutation have equivalent fitness consequences for their bearer~\cite{charlesworth_other_2007}.
Similarly, at the population level, both types of mutations will result in a positive transmission bias of the beneficial allele.
However, at the macro-evolutionary scale, the consequences of these two types of mutations are fundamentally different.
While adaptive mutations promote phenotype diversification (fig.~\ref{fig:fitness-landscape}C), beneficial non-adaptive mutations promote phenotype stability and may help preserve well-established biological systems (fig.~\ref{fig:fitness-landscape}D).
Additionally, the direction of adaptive evolution is unpredictable because it is caused by an unforeseen change in the environment and, hence, in the underlying fitness landscape~\cite{bazykin_changing_2015}.
On the other hand, beneficial non-adaptive mutations are predictable because, under a stable fitness landscape, any change from non-optimal to optimal amino acids will move back the site toward the equilibrium expected under the fitness landscape~\cite{moses_inferring_2009, fischer_germline_2011, chen_hunting_2021}.
They can then be distinguished from truly novel beneficial mutations because the latter are not expected to mutate toward the amino acids of higher fitnesses defined by the stable fitness landscape but rather mutate to amino acids showing a diversified pattern (fig.~\ref{fig:fitness-landscape}).

\begin{figure*}[htb]
       \centering
       \includegraphics[width=17.8cm, page=1]{figure1}
       \caption{
              (A \& B) For a given codon position of a protein-coding DNA sequence, amino acids (x-axis) have different fitness values (y-axis).
              Under a fluctuating fitness landscape (A), these fitnesses change with time.
              The protein sequence follows the moving target defined by the amino-acid fitnesses. Since substitutions are preferentially accepted if they are in the direction of this target, substitutions are, on average, adaptive.
              At the phylogenetic scale (C), beneficial substitutions are common (positive signs), promoting phenotype diversification across species.
              Under a stable fitness landscape (B), most mutations reaching fixation are either slightly deleterious reaching fixation due to drift or are beneficial non-adaptive mutations restoring a more optimal amino acid.
              At the phylogenetic scale (D), deleterious substitutions (negative signs) are often reverted via beneficial non-adaptive mutations (positive signs), promoting phenotype stability and preserving well-established biological systems.
              Even though, individually, any beneficial non-adaptive mutation might have a weak effect on its bearer, we expect them to be scattered across the genome and the genome-wide signature of beneficial non-adaptive mutations to be detectable and quantifiable.}
       \label{fig:fitness-landscape}
\end{figure*}

\begin{figure*}[htb]
       \centering
       \includegraphics[width=17.8cm, page=1]{figure2}
       \caption{
              Selection coefficients at the phylogenetic and population scales.
              At the phylogenetic scale (A), we estimated the amino-acid fitness for each site from protein-coding DNA alignments using mutation-selection codon models.
              For every possible mutation, the difference in amino-acid fitness before and after the mutation allows us to compute the selection coefficient at the phylogenetic scale ($\Sphy$).
              Depending on $\Sphy$ (B) mutations can be predicted as deleterious ($\SphyDel$), nearly-neutral ($\SphyNeu$) or beneficial non-adaptive mutations ($\SphyBen$) toward a fitter amino acid and repairing existing functions.
              At the population scale, each observed single nucleotide polymorphism (SNP) segregating in the population can also be classified according to its $\Sphy$ value (C).
              Occurrence and frequency in the population of non-synonymous polymorphisms, contrasted to synonymous polymorphisms (deemed neutral) is used to estimate selection coefficients (D-E) at the population scale ($\Spop$), for each class of selection ($\SphyDel$, $\SphyNeu$, $\SphyBen$).
              We can thus assess whether $\Sphy$ predicts $\Spop$ and compute \textit{precision} (F) and \textit{recall} (G) for each class.
              The \textit{recall} value for class $\SphyBen$ is the probability for beneficial mutations to be non-adaptive (G).
       }
       \label{fig:method}
\end{figure*}

@@

Theoretically, we can thus use large-scale genomic data to assess whether the fitness estimated at the phylogenetic scale predicts the fitness effect at the population scale for both deleterious and beneficial non-adaptive mutations.
The placental mammals represent an excellent study system to perform such analysis.
Having originated $\sim$102 million years ago, they diversified quickly~\cite{foley_genomic_2023}.
Additionally, polymorphism data are available for many species~\cite{howe_ensembl_2021}, as are high quality protein-coding DNA alignments across the genome~\cite{ranwez_orthomam_2007, scornavacca_orthomam_2019}.
By performing our analysis on 14,509 orthologous protein-coding genes across 87 species, we focus on genes shared across all mammals in our dataset and not newly functionalized genes in a lineage.

For each gene, fitting the mutation-selection model to the multi-species sequence alignment, assuming that the underlying fitness landscape is stable along the phylogenetic tree, allows us to obtain relative fitnesses for all amino acids for each site of the alignment (fig.~\ref{fig:method}A).
Given these relative fitnesses ($F$) for each amino acid, the difference in fitness between a pair of amino acids is the scaled selection coefficient ($\Sphy = \Delta F$), which is formally the product of the selection coefficient at the individual level ($s$) and the effective population size ($\Ne$), as $\Sphy = 4 \Ne \times s$. The value of $\Sphy$ informs us on the strength of selection exerted on amino acids changes.
Thus, according to its $\Sphy$ value, we can classify any mutation as either a deleterious mutation toward a less fit amino acid ($\SphyDel \coloneqq \Sphy < -1$), a nearly-neutral mutation ($\SphyNeu \coloneqq -1 < \Sphy <1$), or a mutation toward a known fitter amino acid constituting thus a beneficial non-adaptive mutation ($\SphyBen \coloneqq \Sphy > 1$).

Having identified which potential DNA changes represent beneficial non-adaptive mutations (fig.~\ref{fig:method}B), we retrieved polymorphism data from 28 wild and domesticated populations belonging to six genera (\textit{Equus}, \textit{Bos}, \textit{Capra}, \textit{Ovis}, \textit{Chlorocebus}, and \textit{Homo}) to assess the presence of beneficial non-adaptive mutations at the population scale.
We focused on mutations currently segregating within populations and substitutions in the terminal branches, and checked if any of these observed changes were indeed beneficial (fig.~\ref{fig:method}C-E).
A similar approach demonstrated the presence of beneficial non-adaptive mutations in humans~\cite{moses_inferring_2009, fischer_germline_2011} and in plants~\cite{chen_hunting_2021}.
However, the model used to reconstruct the static fitness landscape in these studies can only be applied to deeply conserved protein domains in the tree of life, which corresponds to a subpart of the proteome that evolves slowly.
The mutation-selection model used in the present work integrates phylogenetic relationships, and thus allows us to estimate the fitness landscape in shallower phylogenetic trees, and therefore can be applied almost exome-wide~\cite{rodrigue_mutationselection_2010}.

We first quantified the likelihood of any DNA mutation to be a beneficial non-adaptive mutation, that is, whenever a DNA mutation increases fitness under a stable fitness landscape ($\SphyBen$).
Subsequently, by quantifying the total amount of beneficial mutations in the current population across all types of DNA mutations ($\SphyDel$, $\SphyNeu$ and $\SphyBen$), we could tease apart beneficial non-adaptive from adaptive mutations resulting from a change in the fitness landscape.
Altogether, in this study, by integrating large-scale genomic datasets at both phylogenetic and population scales, we propose a way to explicitly quantify the contribution of beneficial non-adaptive mutations to positive selection across the entire exome of the six genera (figs.~\ref{fig:method}F-G).

\section*{Results}

\subsection*{Selection along the terminal branches}
First, we assessed whether fitness effects derived from the mutation-selection model at the phylogenetic scale predict selection occurring in terminal branches.
We recovered the mutations that reached fixation in the terminal branches of the six genera.
We only considered mutations fixed in a population as substitutions in the corresponding branch by discarding mutations segregating in our population samples.
We classified each substitution identified in the terminal branches as either $\SphyDel$, $\SphyNeu$, or $\SphyBen$ depending on its $\Sphy$ value obtained at the phylogenetic scale (fig.~\ref{fig:method}A-C).
Because estimation $\Sphy$ was based on the assumption that the fitness landscape is stable across mammals, $\SphyBen$ mutations (i.e., with $\Sphy>1$) bring the bearer of this mutation toward an amino acid predicted to be fitter across mammals.
Importantly, the mammalian alignment used to estimate the amino acid fitness landscape did not include the six focal genera and their sister species.
This ensures independence between, on the one hand, the fitness landscape estimated, and on the other hand, both substitutions that occurred in the terminal branches, and segregating polymorphisms of the focal populations.
Example substitutions in the terminal lineage of \textit{Chlorocebus sabaeus} which are classified as $\SphyBen$ are shown in supplementary materials (section S2).
For instance, in the mammalian protein-coding DNA alignment of gene \textit{SELE}, the nucleotide at site 1722 has mutated (from T to C) at the base of Simiiformes (monkeys and apes), modifying the corresponding amino acid from Serine to Proline, but has been subsequently reverted in the branch of \textit{Chlorocebus sabaeus} (fig.~S1).
However, other substitutions classified as $\SphyBen$ in the terminal branch of \textit{Chlorocebus sabaeus} cannot be clearly interpreted as reversions along the terminal branch, and show several transitions to this amino acid across the mammalian phylogeny, as for instance site 3145 of gene \textit{THSD7A} (fig.~S2).

Among all the substitutions found in each terminal branch, between 10 and 13\% were $\SphyBen$, while $\SphyBen$ mutations only represent between 0.9 and 1.2\% of all non-synonymous mutations (table~S1 and fig.~\ref{fig:homo-afr-results}A-B for humans).
Furthermore, since in principle, $\SphyBen$ mutations are bound to reach fixation more often than neutral mutations, we calculated the $\dnds$ ratio of non-synonymous over synonymous divergence for all terminal lineages, focusing on the non-synonymous changes predicted as $\SphyBen$ mutations ($\dn(\SphyBen) / \ds$).
We obtained values between 1.17 and 1.75 in the different lineages (table~S2), meaning that $\SphyBen$ mutations reach fixation slightly more frequently than synonymous mutations that are supposed to be neutral, consistent with these $\SphyBen$ mutations being weakly beneficial.

This result further indicates that using $\dnds$ as an estimate of adaptation is biased due to the presence of beneficial non-adaptive mutations among the non-synonymous substitutions.
By discarding all beneficial non-adaptive mutations we can obtain an estimate of $\dnds$ which is not inflated.
By comparing these two ways of calculating $\dnds$ (see section~\ref{subsec:substitution-mapping-in-the-terminal-branch} in Materials \& Methods), we calculated that beneficial non-adaptive mutations inflate $\dnds$ values by between 9 and 12\% across genera. This represents a substantial increase when considering that beneficial non-adaptive mutations only represent between 0.9 and 1.2\% of non-synonymous mutational opportunities (table~S3).

@@

In humans, some SNPs have been associated with specific clinical prognosis terms obtained by clinical evaluation of the impact of variants on human Mendelian disorders~\cite{landrum_clinvar_2018}.
Although this classification also relies on deep protein alignments and therefore cannot be considered an independent result from our own, it does provide a consistency check if the effect of a mutation on human health is in line with its fitness effect predicted by our method.
Therefore, we investigated whether the non-synonymous SNPs classified as $\SphyDel$ or $\SphyBen$ showed enrichment in specific clinical terms compared to SNPs classified as $\SphyNeu$.
Our results show that SNPs predicted as deleterious are associated with clinical terms such as \textit{Likely Pathogenic} and \textit{Pathogenic}, implying that, in general, the selective pressure of a mutation exerted across mammals is also predictive of its clinical effect in humans (table~S4)~\cite{sullivan_leveraging_2023}.
Conversely, $\SphyBen$ mutations are associated with clinical terms such as \textit{Benign} and \textit{Likely Benign}, which shows that $\SphyBen$ mutations are less likely to be functionally damaging (table~S5).

@@

Mutations predicted as $\SphyNeu$ were effectively composed of a mix of neutral and selected mutations with varying \textit{precision} (36--63\%) and \textit{recall} (32--45\%) across the different populations (table~\ref{table:proba}, fig.~\ref{fig:homo-afr-results}D for humans).
The variable proportions between populations can be explained by the effective number of individuals in the population ($\Ne$), a major driver of selection efficacy.
Moreover, estimates of mutation rate per generation ($u$), from~\cite[Bergeron et al]{bergeron_evolution_2023} and~\cite[Orlando et al]{orlando_recalibrating_2013}, and Watterson’s $\theta$ obtained from the synonymous SFS as in~\cite[Achaz]{achaz_frequency_2009} allow us to obtain $\Ne$ through $\Ne=\theta/4 u)$.
Using correlation analyses that accounted for phylogenetic relationship (see section \ref{subsec:correlation-diversity} in Materials \& Methods), we found that higher $\Ne$ was associated with a smaller proportion of nearly-neutral mutations ($r^2=0.31$, $p=0.001$, fig.~\ref{fig:diversity}A).
This result follows the prediction of the nearly-neutral theory and suggests that in populations with higher diversity (e.g.,~\textit{Bos} or \textit{Ovis}), discrimination between beneficial and deleterious mutations is more likely to occur (figs.~S3-S5).
Conversely, many more mutations are effectively neutral in populations with lower diversity (e.g.,~\textit{Homo}).

Finally, mutations predicted to be $\SphyBen$ were indeed beneficial for individuals bearing them, with a \textit{precision} (fig.~\ref{fig:method}F) in the range of 19--87\% (table~\ref{table:proba} and fig.~\ref{fig:homo-afr-results}D for humans).
This result confirms that selection toward amino acids restoring existing functions is ongoing in these populations.
Importantly, the \textit{recall} value in this case, computed as $\proba [ \SphyBen  \given  \SpopBen]$, is the probability for a beneficial mutation at the population scale to be a non-adaptive, i.e., going toward a  fitter amino acid given a stable fitness landscape (fig.~\ref{fig:method}G, table~S6).
In other words, the \textit{recall} value quantifies the number of beneficial mutations restoring damaged genomes instead of creating adaptive innovations.
Across the 28 populations, this proportion is in the range of 11--82\% (table~\ref{table:proba}).
